---
title: "Exploratory Analysis: Product Failure Dataset"
subtitle: "BAN-502 Project Phase 1"
author: "Logan King"
format: revealjs
theme: sky
transition: slide
execute:
  echo: false
  warning: false
  message: false
---

## Exploratory Analysis 
```{r setup}
library(tidyverse)
library(GGally)
library(scales)
library(gt)

train_raw <- read_csv("train.csv") # for sanity check

# Load data
train <- read_csv("train.csv") |>
  mutate(id = as.character(id)) |>
  mutate(across(
    c(
      product_code
    ),
    as_factor
  )) |>
  mutate(across(c(attribute_0, attribute_1), parse_number)) |> 
  mutate(failure = ifelse(failure == "Yes", 1, 0))

```

- 26,570 observations across 26 variables
- Predictor Variable: `failure` (0 = no failure, 1 = failure)
- Attribute 0 and 1 were changed to be numeric for better analysis

::: {.columns}
::: {.column width="40%"}
```{r variable-summary}
# Identify variable types
numeric_vars <- train |>
  select(where(is.numeric)) |>
  names()

categorical_vars <- train |>
  select(where(is.factor), -failure) |>
  names()

# Summary table
tibble(
  `Variable Type` = c(
    "Numeric Predictors",
    "Categorical Predictors",
    "Response Variable"
  ),
  Count = c(length(numeric_vars), length(categorical_vars), 1),
  Examples = c(
    paste(head(numeric_vars, 5), collapse = ", "),
    paste(head(categorical_vars, 4), collapse = ", "),
    "failure (0, 1)"
  )
) |>
  gt() |>
  tab_header(title = "Dataset Variable Summary") |>
  tab_options(
    heading.background.color = "#3fb1e3",
    heading.title.font.size = px(18),
    column_labels.background.color = "#e3f2fd",
    table.background.color = "#f8f9fa",
    table.width = pct(100),
    table.font.size = px(14)
  ) |>
  tab_style(
    style = list(
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_title()
  ) |>
  tab_style(
    style = cell_text(color = "#375a7f", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_label(
    `Variable Type` = "Type",
    Examples = "Sample Variables"
  )
```

:::

::: {.column width="60%"}
```{r failure-distribution, fig.width=8, fig.height=5}
# Calculate summary statistics
failure_summary <- train |>
  count(failure) |>
  mutate(
    pct = n / sum(n) * 100,
    label = sprintf("%s\n%.1f%%", comma(n), pct)
  )

# Create plot
ggplot(train, aes(x = failure)) +
  geom_bar(fill = c("#3fb1e3", "#6ecff6"), alpha = 0.9, width = 0.6) +
  geom_text(
    data = failure_summary,
    aes(y = n, label = label),
    vjust = -0.5,
    size = 5.5,
    color = "#375a7f",
    fontface = "bold",
    lineheight = 0.9
  ) +
  labs(
    title = "Failure Distribution",
    x = "Failure Status",
    y = "Count"
  ) +
  scale_x_discrete(labels = c("0" = "No Failure", "1" = "Failure")) +
  scale_y_continuous(
    limits = c(0, max(failure_summary$n) * 1.15),
    expand = expansion(mult = c(0, 0.15)),
    labels = comma
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(
      color = "#375a7f",
      size = 20,
      face = "bold",
      hjust = 0.5,
      margin = margin(b = 5)
    ),
    plot.subtitle = element_text(
      color = "#6c757d",
      size = 13,
      hjust = 0.5,
      margin = margin(b = 15)
    ),
    axis.title = element_text(color = "#6c757d", size = 13, face = "bold"),
    axis.text = element_text(color = "#495057", size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "#e3f2fd", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#f8f9fa", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(20, 20, 20, 20)
  )
```

:::
:::
## Missing Data Analysis
```{r}
# Calculate missing data
missing_summary <- train |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) |>
  mutate(
    missing_pct = missing_count / nrow(train) * 100,
    has_missing = missing_count > 0
  ) |>
  filter(has_missing) |>
  arrange(desc(missing_pct))

rows_with_missing <- train |>
  filter(if_any(everything(), is.na)) |>
  nrow()

pct_rows_affected <- (rows_with_missing / nrow(train)) * 100
```

::: {.columns}
::: {.column width="60%"}
**Key Findings:**

- `r nrow(missing_summary)` variables with missing data
- Maximum: `r round(max(missing_summary$missing_pct), 1)`%
- **Cumulative impact:** `r comma(rows_with_missing)` rows (`r round(pct_rows_affected, 1)`%) affected
- Row deletion would discard **significant** data so we will have to impute missing values
- No missing values in response variable
:::

::: {.column width="40%"}
```{r}
#| fig-width: 5
#| fig-height: 3.5

missing_summary |>
  ggplot(aes(x = reorder(variable, missing_pct), y = missing_pct)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(
    aes(label = paste0(round(missing_pct, 2), "%")),
    hjust = -0.1,
    size = 3
  ) +
  coord_flip() +
  labs(title = "Missing Data by Variable", x = NULL, y = "% Missing") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_blank()
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

:::
:::
## Strong Predictors: Correlation of Variables

```{r}
train_loading_attr <- train |>
  select(loading, starts_with("attribute_"), failure)

train_measurement <- train |>
  select(starts_with("measurement"), failure)

ggcorr(train)
```

- There is not a very strong correlation between any of the predictors and the response variable, but there are some moderate correlations between loading and failure that could be explored further in Phase 2.

## Phase 2 Recommendations
- **Imputation Strategy**: Given the significant amount of missing data, I recommend using  mean imputation for numeric variables. This will allow us to retain as much data as possible while minimizing bias.
- **Modeling Approach**: Start with a logistic regression model to establish a baseline, then explore more complex models like random forests or boosting  to capture potential nonlinearities and interactions in the data.